Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-20 doc/lean.tex
Overall tuning

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the \lean{} programming language
and proof assistant~\cite{demoura:cade:2021}.
In this section, we briefly describe the functions we have implemented in Lean and
the specifications we have proved.
More information is provided in the Appendix~\ref{appendix:lean}.

>>>>

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both standard and weighted model counters, in the \lean{} programming language
and proof assistant~\cite{demoura:cade:2021}.
In this section, we briefly describe the functions we have implemented in Lean and
the specifications we have proved.
More information is provided in the Appendix~\ref{appendix:lean}.

### 2023-03-20 doc/mc-paper.tex
phi_I
Minor wording changes
Overall tuning

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\phi$}};
  \node [above] at (2,20) {\texttt{.cnf}};

@@

  %% OK/Not
  \draw[thick] (62,14) [-latex] -- (66,14) ;
  \node [right] at (66,15.2) { OK /} ;
  \node [right] at (66,12.8) { Not OK} ;
  \draw[thick] (62,4) [-latex] -- (66,4) ;
  \node [right] at (66,4) {\large {$\textbf{R}(\phi)$}};

@@

Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we 
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can also validate the unweighted or
weighted count generated from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

@@

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $r$ and the defining clauses of the POG nodes. Recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node, although
our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root node $\noder$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
that any total assignment $\assign$ that is a model of $\inputformula$
and the POG definition clauses
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$r$ as a unit clause.  We have devised two methods of generating this
proof.  The {\em monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The {\em structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It too may require calls to a proof-generating SAT
solver.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
We can express the negation of $\phi_\noder$ in clausal form as
$\theta_{\obar{\noder}} = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
when the formula $\inputformula \land \theta_{\obar{\noder}}$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $r$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

The recursive call $\validate(u, \passign, \psi)$ assumes that we have
traversed a path from the root node down to node $\nodeu$, with the
literals encountered in the product nodes forming the partial
assignment $\passign$.  The set of clauses $\psi$ can be a proper
subset of the input clauses $\inputformula$ when a product node has caused
a splitting into clauses containing disjoint variables.
The subgraph with root node $\nodeu$ should be a POG representation of the formula
$\simplify{\psi}{\passign}$.  Having the
routine add its target clause indicates that any total assignment
$\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root node $\noder$ requires showing that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\phi$.  Deleting $C$ requires proving that
$\phi \imply C$.

@@

Working upward through the graph, node $\nodeu$ is {\em marked} when
the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
program adds $\obar{u}$ to the RUP literals and adds the appropriate
defining clause to the hint.  A literal node for
$\lit$ will be marked if $\lit \in C$, with no hint required.  If
product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
added to the hint.  Marking sum node $\nodeu$ requires that its two children are marked.
The first defining
clause for this node (see Table~\ref{tab:defining}(B)) will then be added as a hint.  At the very end, the program
(assuming the reverse implication holds) will attempt to mark root
node $\noder$, which would require $\assign(r) = 0$, yielding a
conflict.

@@

{\em Lemmas:} Our formulation of $\validate$ requires each call at a
node $\nodeu$ to recursively validate all of its children.  This
effectively expands the graph into a tree, potentially requiring an
exponential number of recursive steps.  Instead, for each node
$\nodeu$ having $\indegree(u) > 1$, the program can define and
generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
a call to $\validate$ and then apply this lemma for this and
subsequent calls.  The lemma states that the POG with root node
$\nodeu$ satisfies forward implication for a formula
$\gamma_{\nodeu}$, where some of the clauses in this formula are
input clauses from $\inputformula$, but others are simplified
versions of input clauses.
The key idea is to introduce product nodes to encode (via DeMorgan's
Laws) the simplified clauses and have these serve as lemma arguments.

@@

We found that the monolithic approach for generating the forward
implication proof works well for smaller POGs, but it becomes
unreliable and inefficient for larger ones.  These experiments suggest
a possible hybrid approach, stopping the recursion of the structural
approach and shifting to monolithic mode once a subgraph size is below some
threshold.  We found that our two optimizations: literal grouping and
lemmas can provide substantial improvements in proof size and runtime.

@@

Finally, the runtimes for the MICE toolchain versus our CRAT toolchain
showed little correlation, reflecting the fact that the two solve
slightly different problems and use different approaches.  In general,
our CRAT toolchain showed better scaling, in part due to its ability
to control the recursion through lemmas.

>>>>

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\inputformula$}};
  \node [above] at (2,20) {\texttt{.cnf}};

@@

  %% OK/Not
  \draw[thick] (62,14) [-latex] -- (66,14) ;
  \node [right] at (66,15.2) { OK /} ;
  \node [right] at (66,12.8) { Not OK} ;
  \draw[thick] (62,4) [-latex] -- (66,4) ;
  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};

@@

Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we 
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can also validate the standard or
weighted count generated from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

@@

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
Every POG must have a designated {\em root node} $\noder$.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
any total assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
A d-DNNF
graph can be directly translated  into a POG,
although in doing this conversion,
our program also performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root node $\noder$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
that any total assignment $\assign$ that is a model of $\inputformula$
and the POG definition clauses
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$\{r\}$ as a unit clause.  We have devised two methods of generating this
proof.  The {\em monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The {\em structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It too may require calls to a proof-generating SAT
solver.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
We can express the negation of $\phi_\noder$ in clausal form as
$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
when the formula $\inputformula \land \theta_{\obar{\noder}}$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $\{r\}$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

The recursive call $\validate(u, \passign, \psi)$ assumes that we have
traversed a path from the root node down to node $\nodeu$, with the
literals encountered in the product nodes forming the partial
assignment $\passign$.  The set of clauses $\psi$ can be a proper
subset of the input clauses $\inputformula$ when a product node has caused
a splitting into clauses containing disjoint variables.
The subgraph with root node $\nodeu$ should be a POG representation of the formula
$\simplify{\psi}{\passign}$.
%% Having the
%% routine add its target clause indicates that any total assignment
%% $\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root node $\noder$ requires showing that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.
%%  Suppose at some point in this process that we want to
%% delete clause $C$ such that the remaining set of clauses will form the
%% CNF formula $\phi$.  Deleting $C$ requires proving that
%% $\phi \imply C$.

@@

Working upward through the graph, node $\nodeu$ is {\em marked} when
the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
program adds $\obar{u}$ to the RUP literals and adds the appropriate
defining clause to the hint.  A literal node for
$\lit$ will be marked if $\lit \in C$, with no hint required.  If
product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
added to the hint.  Marking sum node $\nodeu$ requires that its two children are marked.
The first defining
clause for this node (see Table~\ref{tab:defining}(B)) will then be added to the hint.  At the very end, the program
(assuming the reverse implication holds) will attempt to mark root
node $\noder$, which would require $\assign(r) = 0$, yielding a
conflict.

@@

{\em Lemmas:} Our formulation of $\validate$ requires each call at a
node $\nodeu$ to recursively validate all of its children.  This
effectively expands the graph into a tree, potentially requiring an
exponential number of recursive steps.  Instead, for each node
$\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
a call to $\validate$ and then apply this lemma for this and
subsequent calls.  The lemma states that the POG with root node
$\nodeu$ satisfies forward implication for a formula
$\gamma_{\nodeu}$, where some of the clauses in this formula are
input clauses from $\inputformula$, but others are simplified
versions of input clauses.
The key idea is to introduce product nodes to encode (via DeMorgan's
Laws) the simplified clauses and have these serve as lemma arguments.

@@

We found that the monolithic approach for generating the forward
implication proof works well for smaller POGs, but it becomes
unreliable and inefficient for larger ones.  These experiments suggest
a possible hybrid approach, stopping the recursion of the structural
approach and shifting to monolithic mode once the subgraph size is below some
threshold.  We found that our two optimizations: literal grouping and
lemmas can provide substantial improvements in proof size and runtime.

@@

Finally, the runtimes for the MICE toolchain versus our CRAT toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general,
our CRAT toolchain showed better scaling, in part due to its ability
to control the recursion through lemmas.
